Leadership across cultures and contexts: an exploratory analysis

### 2020-05-11 leadership_across_cultures_contexts.Rmd
Revisions to status and dual section of discussion.

Although evidence for high status leaders did not substantially vary by subsistence or group structure, this null result suggests relationships between leadership and social status are not specific to social complexity or substantially context-dependent and if anything leadership among lower levels of social organization and less politically complex societies (i.e., kin groups and hunter-gatherers) is more likely to be associated with high social status, relative to other groups (see \@ref(fig:groupstatus)A and \@ref(fig:groupstatus)B), which could be related to institutionalized and overt or ascribed social distinctions among more politically complex populations. Nonetheless, these findings question the informed speculation by @smith_leadership_2020 that, "leadership may therefore be more strongly correlated with high status in large, complex organizations, such as in corporations and governments (p. 2)." However, our text analytic results, which relied on the semantic context of text records to predict the presence of high status leaders generally identified terms related to status and prestige (i.e., *respect*, *prestige*, *honor*) and possibly dominance or proscribed hierarchy (i.e., *rank*, *position*), as stronger predictors of high status leader than terms associated with leadership per se (i.e., *influence*, *authority*, *headman*), which lends some support to the distinction between status and leadership highlighted by @smith_leadership_2020 (though we caution from drawing strong interpretations from these text-analytic results, see Figure S\@ref(fig:highstatusplot)). Overall, these results suggest associations between leadership and high status are ubiquitous across human societies, including among kin groups and among hunter-gatherers and that while effective leaders across a wide range of social contexts are the recipients of increased social status for their valuable services, ineffective leadership is associated with reduced social status.

@@

Only a small fraction of the ethnographic record sampled here discussed female leaders. Our results do however provide some insight on female leadership and sex-differences in leadership across cultures and contexts. Female leaders were more likely to emerge in kin groups and residential subgroups, consistent with theoretical arguments emphasizing women as leaders of families by @garfield_evolutionary_2019 and @hagen_leadership_2019 and previous cross-cultural studies [e.g., @yanca_female_2004]. Texts describing female leaders exclusively also provided a relatively higher proportion of total leader benefits (Figure \@ref(fig:leadertotalbenefits) and descriptions of high status leaders (Figure \@ref(fig:groupstatus)) compared to descriptions of leaders that were not sex-specific (holding other measures constant; although this effect was similar to benefits to and high status of male-leader texts). This suggests, much like male leaders who are more often political leaders, female leaders are also high status individuals and receive substantial benefits for their services.

Across cultures and contexts, when ethnographic texts described female leaders exclusively, they were relatively more likely to be described as older, of high status, married to a high-quality spouse in a polygynous marriage, and the recipients of various mating benefits. This suite of characteristics is consistent with discussions of high status first-wives who gain and maintain increased social influence across the lifespan by leveraging their extended kin and social networks [@goodale_tiwi_1971; @brown_her_1985], and supports the view of positive assortative mating of high status leaders @garfield_investigating_2019 documented among elected leaders in a small-scale society with relative gender-egalitarianism. Evidence for male leaders (relative to evidence for female leaders or evidence that was not sex-specific) was indeed linked to group-level and political functions and male leaders were relatively more likely to provide social services, control political structure, and were at greater risk of losing their high social standing. Female leaders, relative to males, were also more likely to lack coercive authority, which is consistent with much of leadership studies suggesting dominant and aggressive women are less likely to be preferred as leaders, unlike dominant men who are sometimes preferred as leaders [@garfield_investigating_2019; @williams_subtle_2016]. See Figure S\@ref(fig:heatmapsex).

Despite sex differences, there is also likely to be substantial overlap in many of the qualities and functions of female and male leaders across human societies. Leader sex was not a predictor of the majority of the measures investigated here (adjusting for other population-level measures; see Table S\@ref(tab:universalvars)). When sex was a predictor (Table S\@ref(tab:variablevars)), descriptions including both female and male leaders provided relatively more supporting evidence for many key measures of leader qualities and functions, including group organization, resource distribution, ritual functions, and being brave and experienced and accomplished )Figure S\@ref(fig:heatmapsex).

@@

Leadership and social influence are expected to be associated with particular suites of individual qualities and functions, which potentially covary by cultural or social context, personality, or other inter-individual variation [@cheng_two_2013; @von_rueden_roots_2014; @hogan_personality_2020; @lukaszewski_extraversion_2015; @henrich_evolution_2001; @bass_handbook_1990]. Our systematic cross-cultural evidence supports many aspects of influential theories. Much of the evolutionary theoretical literature on leadership, however, has suggested social influence is most often achieved via one of two distinct strategies: prestige or dominance. The dominance-prestige theory as formulated by @henrich_evolution_2001 significantly advanced evolutionary approaches to leadership and social hierarchy and inspired much subsequent work across the behavioral sciences [e.g., @van_vugt_dual_2019; @cheng_two_2013; @maner_j.k_dominance_2016; @halevy_status_2012; @suessenbach_dominance_2019; @maner_dominance_2017; @garfield_evolutionary_2019]. @van_vugt_dual_2019 [p. 1], for example, "argue that there are two leadership styles: prestige style and dominance style," and @cheng_two_2013 [p. 119] in focusing on dominance and prestige as distinct leadership strategies suggest that, "human social stratification is characterized by the co-occurrence of both strategies." Exploratory cluster analyses reported here, however, suggest multiple features of leader qualities and leader functions across cultures and contexts, beyond dominance and prestige. Leaders likely rely a range of individual capacities, including physical, supernatural, and cognitive endowments, as well as their social networks, in order to organize group members, implement strategic actions, and provide prosocial services to the group, all while conforming to cultural pressures. Compartmentalizing this range of diversity into two distinct social strategies, although theoretically and empirically convenient, likely misses much of the ethnographic reality of human leadership dynamics.

<!-- DISCUSSES PROSOCIALITY FEATURE -->
The prosociality feature of leader qualities, including generosity, interpersonal and oratory skills, and charisma as well as the role of cultural values shaping leadership styles, including proper behavior and having a favorable personality, generally correspond to a charismatic style of leadership that has been identified in industrialized settings and widely discussed in sociology and more recently emphasized in evolutionary psychology models of leadership [@den_hartog_culture_1999; @weber_distribution_1978; @grabo_charismatic_2016; @howell_role_2005]. The importance of prosociality  also corresponds to @henrich_big_2015's Big-man mechanism, which suggests positive assortment of followers to prosocial leaders can facilitate the evolution of cooperative social norms and prestige-based leadership.
<!-- A few variables did not seem to fit prosocial theme, however, such as *Physical health*, *Physical formidable*, and *Drug consumption* (although drug use is often an important social activity). -->

@@

<!-- DISCUSSES COMPETENCIES FEATURE -->
The individual-level competencies was associated with variables associated with dominance-based leadership, including coercive authority and aggressiveness [@tiger_imperial_1971; @barkow_prestige_1975; @barkow_darwin_1989; @henrich_evolution_2001; @cheng_two_2013] as well as a grouping that included qualities of knowledgeable and intelligent and experienced and accomplished, traits that have been emphasized by the *Prestige* theory [@henrich_evolution_2001; @cheng_two_2013] and the more general *Computational services* theory [@hagen_leadership_2019; @garfield_evolutionary_2019], and which are among the most strongly supported in the ethnographic record (Figure \@ref(fig:varsupport)), a point to which we shall return. Interestingly, being feared and having supernatural qualities clustered more closely with the prestige-style traits than with the dominance-style traits. There is likely a strong connection between shamanism and ritual-based influence with both prestige and dominance styles of leadership. Ritual and supernatural capacities may represent a dimension of leader qualities in which dominance and prestige are coupled forming an alternative style of leadership, as shamans often provide a necessary and valued public good, have special expertise and abilities, are feared and achieve or maintain influence via threats or use of direct and/or indirect aggression. Artistic performance is also implicated in this style of leadership and ceremonial, artistic performances by shamans are common displays which often incorporate the superhuman abilities of shamans [@singh_cultural_2017].

As brief examples of the very many in the ethnographic record, among the Azande, "though the most knowledgeable witch-doctors tend to become leaders, there is no institutionalized organization or leadership among them" [@fo07-056, p. 85]; and among the Highland Scots, @es10-003 [p. 147] notes, "on the whole, their long ties with place and people, coupled with education, made them useful leaders of a community in which leaders were needed to adapt life to a changing economy." Such exceptionally talented leaders, however, also face unique challenges, particularly within more egalitarian societies which emphasizes personal autonomy and resit hierarchy in the context of cultural transitions. @nm09-056 [p. 183] describes Laura Cornelius Kellogg (also known as Minnie and Wynnogene, 1880-1947), a land-claims activist, writer, and perhaps the most famous female leader among the Iroquois:

>>>>

Although evidence for high status leaders did not substantially vary by subsistence or group structure, this null result suggests relationships between leadership and social status are not substantially context-dependent or specific to greater social complexity. Speculatively, if anything leadership among lower levels of social organization and less politically complex societies (i.e., kin groups and hunter-gatherers) is more likely to be associated with high social status, relative to other groups (see \@ref(fig:groupstatus)A and \@ref(fig:groupstatus)B). If so, this trend could be due to the institutionalized and overt or ascribed social distinctions among more politically complex populations and the importance of achieve statuses among less politically and socially complex groups. Nonetheless, these findings question the informed speculation by @smith_leadership_2020 that, "leadership may therefore be more strongly correlated with high status in large, complex organizations, such as in corporations and governments (p. 2)." However, our text analytic results, which relied on the semantic context of text records to predict the presence of high status leaders generally identified terms related to status and prestige (i.e., *respect*, *prestige*, *honor*) and possibly dominance or proscribed hierarchy (i.e., *rank*, *position*), as stronger predictors of high status leader than terms associated with leadership per se (i.e., *influence*, *authority*, *headman*), which lends some support to the distinction between status and leadership highlighted by @smith_leadership_2020 (though we caution from drawing strong interpretations from these text-analytic results, see Figure S\@ref(fig:highstatusplot)). Overall, these results suggest associations between leadership and high status are ubiquitous across human societies, including among kin groups and among hunter-gatherers and that while effective leaders across a wide range of social contexts are the recipients of increased social status for their valuable services, ineffective leadership is associated with reduced social status.

@@

Only a small fraction of the ethnographic record sampled here discussed female leaders. Our results do however provide some insight on female leadership and sex-differences in leadership across cultures and contexts (though we emphsize the limitations of our data). Female leaders were more likely to emerge in kin groups and residential subgroups, consistent with theoretical arguments emphasizing women as leaders of families  [e.g., @garfield_evolutionary_2019; @hagen_leadership_2019] and previous cross-cultural studies [e.g., @yanca_female_2004]. Texts describing female leaders exclusively also provided a relatively higher proportion of total leader benefits (Figure \@ref(fig:leadertotalbenefits)) and descriptions of high status leaders (Figure \@ref(fig:groupstatus)) compared to descriptions of leaders that were not sex-specific (holding other measures constant; although this effect was similar to benefits to and high status of male-leader texts). This suggests, much like male leaders who are more often political leaders, female leaders are also high status individuals and receive substantial benefits for their services.

Across cultures and contexts, when ethnographic texts described female leaders exclusively, they were relatively more likely to be described as married to a high-quality spouse in a polygynous marriage, and the recipients of various mating benefits (Figure S\@ref(fig:heatmapsex)). This suite of characteristics is consistent with discussions of high status first-wives who gain and maintain increased social influence across the lifespan by leveraging their extended kin and social networks [@goodale_tiwi_1971; @brown_her_1985], and supports the view of positive assortative mating of high status leaders @garfield_investigating_2019 documented among elected leaders in a small-scale society with relative gender-egalitarianism. Evidence for male leaders (relative to evidence for female leaders or evidence that was not sex-specific) was indeed linked to group-level and political functions and male leaders were relatively more likely to provide social services, control political structure, and were at greater risk of losing their high social standing. Female leaders, relative to males, were also more likely to lack coercive authority, which is consistent with much of leadership studies suggesting dominant and aggressive women are less likely to be preferred as leaders, unlike dominant men who are sometimes preferred as leaders [@garfield_investigating_2019; @williams_subtle_2016]. See Figure S\@ref(fig:heatmapsex).

Despite sex differences, there is also likely to be substantial overlap in many of the qualities and functions of female and male leaders across human societies. Leader sex was not a predictor of the majority of the measures investigated here (adjusting for other factors; see Table S\@ref(tab:universalvars)). When sex was a predictor (Table S\@ref(tab:variablevars)), descriptions including both female and male leaders provided relatively more supporting evidence for many key measures of leader qualities and functions, including group organization, resource distribution, ritual functions, and being brave and experienced and accomplished (see Figure S\@ref(fig:heatmapsex)).

@@

Leadership and social influence are expected to be associated with particular suites of individual qualities and functions, which potentially covary by cultural or social context, personality, or other inter-individual variation [@cheng_two_2013; @von_rueden_roots_2014; @hogan_personality_2020; @lukaszewski_extraversion_2015; @henrich_evolution_2001; @bass_handbook_1990]. Our systematic cross-cultural results support many aspects of influential theories. Much of the evolutionary theoretical literature on leadership, however, has suggested social influence is most often achieved via one of two distinct strategies: dominance or prestige [for review see @garfield_review_2019; @garfield_evolutionary_2019]. The dominance-prestige theory as formulated by @henrich_evolution_2001 significantly advanced evolutionary approaches to leadership and social hierarchy and inspired much subsequent work across the behavioral sciences [e.g., @van_vugt_dual_2019; @cheng_two_2013; @maner_j.k_dominance_2016; @halevy_status_2012; @suessenbach_dominance_2019; @maner_dominance_2017; @garfield_evolutionary_2019; @garfield_investigating_2019]. @van_vugt_dual_2019 [p. 1], for example, "argue that there are two leadership styles: prestige style and dominance style," and @cheng_two_2013 [p. 119] in focusing on distinctions between dominance and prestige suggests, "human social stratification is characterized by the co-occurrence of both strategies." Exploratory cluster analyses reported here, however, suggest multiple features of leader qualities and leader functions, beyond dominance and prestige. Leaders likely rely a range of individual capacities, including physical, supernatural, and cognitive endowments, as well as their social networks, in order to organize group members, implement strategic actions, and provide prosocial services to the group, all while conforming to cultural norms. Dichotomizing this complexity into two distinct social strategies, although theoretically and empirically convenient, likely failes to capture much of the ethnographic reality of human leadership dynamics.

<!-- DISCUSSES PROSOCIALITY FEATURE -->
The prosociality feature of leader qualities, including generosity, interpersonal and oratory skills, and charisma as well as the role of cultural values shaping leadership styles, including proper behavior and having a favorable personality, generally correspond to a charismatic style of leadership that has been identified in industrialized settings and widely discussed in sociology and more recently emphasized in evolutionary psychology models of leadership [@den_hartog_culture_1999; @weber_distribution_1978; @grabo_charismatic_2016; @howell_role_2005]. The importance of prosociality  also corresponds to the @henrich_big_2015 Big-man mechanism, which suggests positive assortment of followers to prosocial leaders can facilitate the evolution of cooperative social norms and prestige-based leadership.
<!-- A few variables did not seem to fit prosocial theme, however, such as *Physical health*, *Physical formidable*, and *Drug consumption* (although drug use is often an important social activity). -->

@@

<!-- DISCUSSES COMPETENCIES FEATURE -->
The individual-level competencies feature was comprised of variables associated with dominance-based leadership, including coercive authority and aggressiveness [@tiger_imperial_1971; @barkow_prestige_1975; @barkow_darwin_1989; @henrich_evolution_2001; @cheng_two_2013] as well as a group that included qualities of knowledgeable and intelligent and experienced and accomplished, traits that have been emphasized by the prestige theory [@henrich_evolution_2001; @cheng_two_2013] and the more general computational services theory [@hagen_leadership_2019; @garfield_evolutionary_2019], and which are among the most strongly supported in the ethnographic record (Figure \@ref(fig:varsupport)). Interestingly, being feared and having supernatural qualities clustered more closely with the prestige-style traits than with the dominance-style traits. There is likely a strong connection between shamanism and ritual-based influence with both prestige and dominance styles of leadership. Ritual and supernatural capacities may represent a dimension of leader qualities in which dominance and prestige are coupled forming an alternative style of leadership, as shamans often provide a necessary and valued public good, have special expertise and abilities, are feared and achieve or maintain influence via threats or use of direct and/or indirect aggression. Artistic performance is also implicated in this style of leadership and ceremonial, artistic performances by shamans are common displays which often incorporate the superhuman abilities of shamans [@singh_cultural_2017].

As brief examples of the diversity of leader capacities, of the very many in the ethnographic record, among the Azande, "though the most knowledgeable witch-doctors tend to become leaders, there is no institutionalized organization or leadership among them" [@fo07-056, p. 85]; and among the Highland Scots, @es10-003 [p. 147] notes, "on the whole, their long ties with place and people, coupled with education, made them useful leaders of a community in which leaders were needed to adapt life to a changing economy." Such exceptionally talented leaders, however, also face unique challenges, particularly within more egalitarian societies which emphasizes personal autonomy and resit hierarchy in the context of cultural transitions. @nm09-056 [p. 183] describes Laura Cornelius Kellogg (also known as Minnie and Wynnogene, 1880-1947), a land-claims activist, writer, and perhaps the most famous female leader among the Iroquois:
